\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2020-12-23 manuscript.Rnw
Some updates to the discussion

%% -- Discussion ---------------------------------------------------------------
\section{Discussion} \label{sec:discussion}
This article provides a formal description of the economic models and cost-effectiveness framework used by the \pkg{hesim} package. An illustrative example is provided for the common three state model used to evaluate treatments in oncology. However, since \pkg{hesim} aims to be flexible, the example only covers a subset of the relevant modeling approaches and functionality. Additional examples as well as documentation for all functions and classes are consequently provided on the package website (\url{https://hesim-dev.github.io/hesim/}).

@@

Alternative approaches might be appropriate in other contexts. For example, if evenly spaced panel data is available, a multinomial logistic regressions can be used to predict transition probabilities. On the other hand, construction of the transition probability matrix and relative treatment effects may not be as straightforward. In these cases, it may be easier to use a \code{define_model()} block to construct transition probability matrices using nonstandard evaluation. When using this approach, each possible combination of the parameters and input data (treatment strategies and patients) is automatically constructed and the user must only write mathematical expressions that use the parameters and input data to create the transition probability matrices. The same \code{define_model()} block may also be used to construct utility and cost models, if desired. Example that leverage both multinomial logistic regession and \code{define_model()} are provided on the course website and the latter is used to reproduce two examples from the reference textbook by \citet{briggs2006decision}.

In Section~\ref{sec:ictstm-example}, we fit a clock-reset multi-state model and simulated outcomes with an individual-level simulation. An iCTSTM can, however, also be developed without fitting a statistical model using \proglang{R} by creating a \code{params_surv} object for each possible transition. If IPD is available for the reference treatment, \code{create_params()} can be use to obtain the baseline \code{params_surv} object from the fitted multi-state model, and relative treatment effects (e.g., hazard ratios) could be added as additional columns to the coefficients stored in the \code{params_surv} object. A similar approach was used in the partitioned survival analysis example above (Section~\ref{sec:psm-example}). A re-analysis of an time-inhomogeneous Markov cohort model of total hip replacement from the \citet{briggs2006decision} textbook is provided on the package website to illustrate.

A third additional feature worth mentioning is that a PSM can be extended to N-states. The 3-state oncology case might, for instance, be extended to account for adverse events or to multiple treatment lines. One related application is the quality-adjusted time without symptoms or toxicity (Q-TWiST) method \citep{goldhirsch1989costs, lenderking1994eval}, which partitions patients into time (i) having toxic side effects, (ii) without symptoms or toxicities, and (iii) post-progression. An example 4-state model is provided on the package website.

As mentioned in the introduction, \code{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. It differs from \pkg{hesim} in that it focuses solely on cDTSTMs (i.e., Markov cohort models) and is not designed to facilitate integration of economic models with the statistical models used for parameter estimation. Another important difference is that \code{hesim} is considerably faster since all simulations are vectorized (i.e, performed at the \proglang{C++} level). Performance tests are provided on the package website. The time-inhomogeneous total hip replacement \citet{briggs2006decision} example was run with both packages using 1,000 PSA iterations. A cohort model in \pkg{heemod} completed in approximately 85 seconds while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \code{hesim} ran in 9 seconds.

There are a number of \proglang{R} multi-state modeling packages that can both fit models and make predictions. \pkg{hesim} makes use of the \pkg{flexsurv} and \pkg{msm} packages for parameter estimation with parametric models A key difference is that \pkg{hesim} is designed specifically for health-economic modeling whereas the other packages are more general purpose statistical packages.

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models. It supports prediction of state probabilities and simulation of multi-state trajectories (for clock-reset models) from cumulative hazard functions evaluated at specific times. \code{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health-economic modeling since it was not designed for it. For instance, the core function for simulating multi-state models, \code{mstate::mssample()}, does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also computationally inefficient for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions where efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran 100 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. The model completed in 0.44 second in \pkg{hesim} and took 34 minutes in \pkg{mstate}, showing that computational considerations are not just an academic concern.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update the input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patients transitions to a new state. Second, we aim to continue to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.

>>>>

%% -- Discussion ---------------------------------------------------------------
\section{Discussion} \label{sec:discussion}
This paper formally describes the economic models and cost-effectiveness framework used by the \pkg{hesim} package. An illustrative example is provided for a three state model commonly used to evaluate treatments in oncology. However, since \pkg{hesim} aims to be flexible, the example only covers a subset of the relevant modeling approaches and functionality. Additional examples as well as documentation for all functions and classes are consequently provided on the package website (\url{https://hesim-dev.github.io/hesim/}).

@@

Alternative approaches might be appropriate in other contexts. For example, if evenly spaced panel data is available, a multinomial logistic regressions can be used to predict transition probabilities. On the other hand, some users may prefer a functional approach that is arguably more human readable. To facilitate this type of approach, a \code{define_model()} block can be used to construct transition probability matrices using nonstandard evaluation. Each possible combination of the parameters and input data (treatment strategies and patients) is automatically constructed and the user can write custom expressions that create the transition probability matrices as a function of the parameters and input data. The same \code{define_model()} block may also be used to construct utility and cost models, if desired. Example that leverage multinomial logistic regression and \code{define_model()} are provided on the package website. The latter is used to reproduce two examples (a simple Markov model of an HIV treatment and a time-inhomogeneous Markov cohort model of total hip replacement) from the reference textbook by \citet{briggs2006decision}.

In Section~\ref{sec:ictstm-example}, we fit a clock-reset multi-state model and simulated outcomes with an individual-level simulation. An iCTSTM can, however, also be developed without fitting a statistical model using \proglang{R} by manually inputting the arguments (e.g., coefficients, probability distributions) of \code{params_surv} objects for each possible transition. If IPD is available for the reference treatment, \code{create_params()} can be use to obtain the baseline \code{params_surv} object from a fitted multi-state model, and relative treatment effects (e.g., hazard ratios) could be added as additional columns to the coefficients stored in the \code{params_surv} object. A similar approach was used in the partitioned survival analysis example above (Section~\ref{sec:psm-example}). A re-analysis of the total hip replacement model from the \citet{briggs2006decision} textbook using an individual-levels simulation is provided on the package website to illustrate.

A third feature worth mentioning is that a PSM can be extended to $N$ health states. The 3-state oncology case might, for instance, be extended to account for adverse events or multiple lines of therapy. One related application is the quality-adjusted time without symptoms or toxicity (Q-TWiST) method \citep{goldhirsch1989costs, lenderking1994eval}, which partitions patients into time spent (i) having toxic side effects, (ii) without symptoms or toxicities, and (iii) after progression. An example 4-state model is provided on the package website.

As mentioned in the introduction, \code{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. It differs from \pkg{hesim} in that it focuses solely on cDTSTMs (i.e., Markov cohort models) and is not designed to facilitate integration of economic models with the statistical models used for parameter estimation. Another important difference is that \code{hesim} is considerably faster since all simulations are vectorized (i.e, performed at the \proglang{C++} level). Performance comparisons are provided on the package website. The time-inhomogeneous total hip replacement \citet{briggs2006decision} example was run with both packages using 1,000 PSA iterations. A cohort model in \pkg{heemod} completed in approximately 85 seconds while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \code{hesim} ran in 9 seconds.

There are a number of multi-state modeling \proglang{R} packages that can both fit models and make predictions. \pkg{hesim} leverages some of these for parameters estimation: \pkg{flexsurv} is used to fit parametric and flexible parametric survival and multi-state models when exact event times are available while \pkg{msm} is used to fit multi-state models when only panel data is available. A key difference is that \pkg{hesim} is designed specifically for health-economic modeling whereas the other packages are more general purpose statistical packages.

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models. It supports prediction of state probabilities and simulation of multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health-economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also computationally inefficient for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions where efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran 100 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. The model completed in 0.44 second in \pkg{hesim} and took 34 minutes in \pkg{mstate}, showing that computational considerations are not just an academic concern.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patient transitions to a new state. Second, we would like to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking in this area, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.
